Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-09 boettiger.tex
alan hastings edits to intro

\section{Introduction}
There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.

We discuss the limitations of current approaches that rely on summary statistics
and introduce a model based approach to detect early warning signals.
We illustrate how stochastic differential equation (SDE) models can be used 
to reflect the uncertainty inherent in the detection of early warning signals.
We caution against paradigms that are not useful for capturing this uncertainty in 
a model-selection based approach, such as information criteria,
and suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets 
and different early warning indicators have in detecting early warning signals.

>>>>

\section{Introduction}
There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs.

A number of particular systems have demonstrated the kinds of relationships that would produce regime shifts, including dynamics of coral reefs (MUMBY ET AL, NATURE), and simple models of metapopulations with differing local population sizes (HASTINGS Hastings, A. (1991). Structured models of metapopulation dynamics. Biological Journal of the Linnean Society, 42, 57-71.).  In cases like these one sensible approach to understanding whether a regime shift would be likely would be to fit the model using either a time series or else independent estimates of spratters. More generally, with a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009}.

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009} that do not depend on estimating the parameters of a known model of a specific system. This has lead to a variety of approaches based on summary statistics (REFS -- ) that look for generic signs of impending regime shifts.  Our goal here is to extend earlier work by providing much more precise estimates of the ability of different statistics to signal impending regime shifts and to develop new approaches that both are more efficient and also lay bare some of the important assumptions underlying attempts to find general warning signs of regime shifts.  We caution against paradigms that are not useful for capturing  uncertainty in
a model-selection based approach, such as information criteria.

We begin our approach by first discussing the limitations of current approaches that rely on summary statistics
and provide a clear description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals.
Finally we suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets
and different early warning indicators have in detecting early warning signals and use this to explore a number of examples.
